Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-16 doc/mc-paper.tex
Polished technical presentation

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}
\newcommand{\validate}{\textsf{validate}}
\newcommand{\prov}{\textrm{Prov}}
\newcommand{\inputformula}{\phi_I}
\newcommand{\pogformula}{\phi_P}

@@

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

@@

Model counting is a challenging problem---more challenging that the
already NP-hard problem of Boolean (un)satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal becomes model counting and not just
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters have developed that scale to very large formulas, as
witnessed by their progress in the annual model counting competitions.

@@

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant.  Running these two programs
has the effect of formally verifying both the logical equivalence of the POG to the input formula and the correctness of the count.
In addition, the
existence of a formally verified checker serves the important role of
establishing the soundness of our proof framework.

@@

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,wetzler14_drattrim,Tan:2021}.  Some progress has also
been made on extending these proof systems to quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.  There tool
chain can be applied to some programs that are pure model counters,
while ours can be applied to pure knowledge compilers.  Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Section~\ref{sect:experimental}.  Our
experience with developing a formally verified CRAT proof checker has shown
us that, even for the well understood framework of extended
resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.

@@

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em negation normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
  negation normal form, and 2) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses in a formula and the
  literals in a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{itemize}
      \item The arguments to a product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a sum must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

We propose a general framework for summarizing properties of Boolean
formulas inspired by the formulation of {\em algebraic model counting}
by Kimmig, et al.~\cite{kimmig:jal:2017}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi, w)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\rep\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\rep\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
As can be seen, a POG is simply a way to represent a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

@@

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step maintaining logical equivalence to
the original formula.  More precisely, the status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables}.  Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
we want to maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  We will use
this form only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  The other is by {\em
  implication redundancy}.  That is, we can add clause $C$ to the set
of clauses forming formula $\phi$ when $\phi \imply C$.  Clauses can
also be deleted through implication redundancy.  That is, given a set
of clauses $\{C\} \cup \phi$, we can remove $C$ if $\phi \imply C$.

@@

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the logical negation of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that any clause identifier $C'$ given in the hints when adding clause $C$ must have $C' < C$.

@@

Finally, the literal denoting the root of the POG is declared with the
{\tt r} command.  It can occur anywhere in the file.  Typically, it
will be the extension variable representing the root of a graph, but in
degenerate cases it can be the positive or negative literal of an
input variable.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG\@.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item There should exactly one remaining clause that was added via RUP
  addition, and this should be a unit clause consisting of the root literal $r$.
\item All of the input clauses have been deleted.
\end{itemize}
The sequence of clause addition steps provide a {\em forward implication} proof that
$\modelset(\phi) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\phi)$.

@@

Figure~\ref{fig:chain} illustrates one method of generating a CRAT
file from an input formula in conjunctive normal form.  Our tool chain
is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
but it would also work for other tools that use a top-down search
strategy to generate a d-DNNF representation.  Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except for the degenerate cases
where the formula is unsatisfiable or a tautology, we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation operators will
only be applied to variables, and so we can view the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG, given by its set of defining clauses $\pogformula$.
Furthermore, it can
negate the POG formula by simply negating its root literal.  Forward implication
will hold
when $\inputformula \imply (r \land \pogformula)$.  Equivalently,
the formula $\inputformula \land \simplify{\pogformula}{\{\obar{r}\}}$
should be unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file, invokes a proof-generating solver, and
imports the proof steps as assertion steps in the CRAT file.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

>>>>

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}
\newcommand{\validate}{\textsf{validate}}
\newcommand{\prov}{\textrm{Prov}}
\newcommand{\inputformula}{\phi_I}
\newcommand{\pogformula}{\theta_P}

@@

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10003794</concept_id>
       <concept_desc>Theory of computation~Automated reasoning</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Automated reasoning}

%\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
\keywords{Propositional model counting, Proof checking}

@@

Model counting is a challenging problem---more challenging that the
already NP-hard problem of Boolean satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal becomes model counting and not just
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters have developed that scale to very large formulas, as
witnessed by their progress in the annual model counting competitions.

@@

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant.  Running these two programs
has the effect of formally verifying both the logical equivalence of
the POG to the input formula and the correctness of the count.  In
addition, the existence of a formally verified checker serves the
important role of establishing the soundness of our proof framework.
Indeed, in the course of developing a formally verified checker, we
identified subtle aspects of the proof obligations required to assure
soundness.

@@

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress has also
been made on extending these proof systems to quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.  Their tool
chain can be applied to some programs that are pure model counters,
while ours can be applied to pure knowledge compilers.  Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Section~\ref{sect:experimental}.  Our
experience with developing a formally verified proof checker has shown
us that, even for the well understood framework of extended
resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.

@@

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em negation normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
  negation normal form, and 2) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to a product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a sum must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

We propose a general framework for summarizing properties of Boolean
formulas inspired by the formulation of {\em algebraic model counting}.~\cite{kimmig:jal:2017}.

@@

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ 
yields the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
{\Large \rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
{\Large \rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
There is a unique {\em root node} $\noder$ such that $\indegree(\noder) = 0$.
As can be seen, a POG is represents a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

@@

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  As we have seen, ring evaluation can readily handle negation.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Although not required for representing d-DNNF graphs,
  both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step maintaining logical equivalence to
the original formula.  More precisely, the status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
we want to maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  The other is by {\em
  implication redundancy}.  That is, we can add clause $C$ to the set
of clauses forming formula $\phi$ when $\phi \imply C$.  Clauses can
also be deleted through implication redundancy.  That is, given a set
of clauses $\{C\} \cup \phi$, we can remove $C$ if $\phi \imply C$.

@@

The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{\nodeu}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
%% The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $\lit$
is represented by a signed integer, with $-v$ being the logical negation of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that any clause identifier $C'$ given in the hints when adding clause $C$ must have $C' < C$.

@@

Finally, the literal denoting the root of the POG is declared with the
{\tt r} command.  It can occur anywhere in the file.  Except in degenerate cases, it
will be the extension variable representing the root of a graph
%, but in
%degenerate cases it can be the positive or negative literal of an
%input variable.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{\nodeu}$.  We write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables.
The sequence of clause addition steps provide a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\inputformula)$.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our tool chain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except for the degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\pogformula$, consisting of the defining clauses of its nodes plus a unit clause for root literal $r$.
Furthermore, it can
negate the POG formula by simply negating its root literal.  More precisely $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{\nodeu}}{\{\obar{r}\}}$
Forward implication
will hold
when $\inputformula \imply \pogformula$.  Equivalently,
the formula $\inputformula \land \neg \pogformula$
should be unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

We describe the structural approach to proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the {\em target} clause:
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add unit clause $r$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.
